Embracing the Laws of Physics: \\ Reversible Models of Computation

### 2017-11-24 survey/survey.tex
continuing with the preliminaries. Did actually find a bunch of text that could be used.

%% Cyber-physical systems increasingly common. Are computing foundations
%% adequate?
%% \jc{your questioning of MoC below is not about CPS! Should either
%% foreshadow the next section 'properly' here, or add a leading
%% example based on CPS to MoC section}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% JC: I am commenting this out as, intereting as this all is, a lot of
%     it is about ``real world'' and NOT about the intersection of physics
%     and computation.  It would be possible to write about that too.
%     For example, the hardware failures of petabyte scale computing is
%     physical; physics-based attacks on privacy and security; physics
%     of dealing with noisy environments (key debouncing, timing drifts,
%     spurious data, heart rate spikes, etc);
\begin{comment}
\section{Physics and computation}

%%%
\subsection{Software: Foundations and Applications}

We begin with a bold thesis: \textit{common computational models
embody various physical assumptions that are plainly
\emph{wrong}}. We certainly
are not the first to make such a statement. We do want to point
out that these physically incorrect computational abstractions
inhibit advances in computing in general.

To appreciate this point, we examine various textbook abstractions and
contrast them with the actual (i.e., physical) computational situations to
which they correspond. Generally speaking, we find that textbook algorithms
differ from ``real programs'' in their assumptions about their environment of
operation. In textbook abstractions, the surrounding environment which
includes, for example, the data on which the algorithm operates is considered
static and well-defined. In the real world, this environment is dynamic and
noisy.

\paragraph*{Sorting.} 
One might consider sorting to be a fairly simple ``solved'' problem. Indeed
textbook sorting algorithms have been formally specified, proved correct,
analyzed in the best case, worst case, average case, and are even taught in
many high school curricula. A beginner programmer can readily implement such
algorithms to sort ``real data'' --- or can she?

Under some assumptions, the answer is clearly yes. However, a moment's
reflection should convince us that the textbook algorithm is too idealized
and that sorting real data in the real world is still a challenge!  Consider
the following quote posted on the Google research blog in 2011:
\begin{quote}
  Almost three years ago we announced results of the first ever ``petasort''
  \ldots . It completed in just over six hours on 4000 computers. Recently we
  repeated the experiment using 8000 computers. The execution time was 33
  minutes, an order of magnitude improvement~\cite{petasort}.
\end{quote}
What happens at the petabyte scale that makes sorting such a
challenge? The answer is simply that ``the real world happens!'' In
the real world, many of the idealized mathematical assumptions
embodied in our computational models are invalidated. For example,
proper hardware functioning can no longer be taken for granted, as
component failure becomes the norm, global synchronization becomes
impossible, determinism becomes expensive to maintain, speculative
execution becomes necessary, global consistency becomes impractical to
maintain, and eventual consistency is accepted; etc. There have been
many attempts to taken some of these issues into account in various
computational models but these attempts are arguably not going far
enough by not questioning the physical assumptions in our
computational models and re-thinking our very foundations.

As another data point to reinforce this point: What algorithm
does Python (and other mainstream platforms) use? A variant of
QuickSort or HeapSort perhaps.  No. Timsort. This hybrid algorithm
performs rather well in practice -- but is complex, and even harder 
to formally analyze. It does not tend to get mentioned in algorithms
textbooks at all.

\paragraph*{Binary Search.} 
In the idealized textbook setting, the binary search algorithm has the
precondition that the data is sorted. Even if we just have a few thousands
items, na\"{\i}vely checking sortedness (in linear time) defeats the purpose of
the efficient (logarithmic) binary search. If we have petabytes of
(changing!) data, is it even possible, in principle, to check sortedness? 
Or guarantee sortedness? In
fact, in this dynamic situation, one is likely to get inconsistent
results. In other words, the very idea of pre-conditions and post-conditions
that can be checked statically appears to be at odds with computing in the
``real world.''

\paragraph*{Security.} 
Proving the security properties of an abstract mathematical model is great
(and necessary). This can be used to conclude that an isolated device is
secure but it is, however, not that relevant for applications based on RFID
devices which are targeted via DPA (Differential Power Analysis)~\cite{dpa}
and and DEMA (Differential Electromagnetic Analysis) attacks. There are
billions of such devices with applications like electronic passports. Power
consumption and electromagnetic properties are abstracted away in our
computational model.

\paragraph*{Privacy.} 
The Internet has created major concerns about privacy. Differential
privacy~\cite{dwork:differential} states that one can achieve privacy
by perturbing the output of a deterministic program by a suitable
amount of noise. The very idea of noise is abstracted away in our
computational and communication models and the amount of noise is a
non-existent abstraction in our typical computational models.

\paragraph*{A Crisis?}
Is this tension between theory and practice a crisis?  We argue:
yes. Computer science theory is founded on abstractions of the real
world. These abstractions are necessary to shield us from details of
technology, noise, errors, etc. and they have served us well until
recently.  It is, however, becoming more and more evident that
computer science theory is becoming less and less relevant to
practice, and hence it is important to revisit our abstractions from a
foundational point of view.
\end{comment}

\section{Models of Computation}
%%%
\begin{comment}
\subsection{Our Origins}

Until Alan Turing's seminal work in 1936, computation was an activity
of the human mind; much like love and peace. While aspects of the idea
predated the twentieth century, it was Turing who established the idea
that computation has a formal interpretation and that all
computability can be captured within a formal system. Implicit in this
achievement however is the idea that abstract models of computation
are just that -- \emph{abstractions of computation realized in the
  physical world.}  In fact, one of the major achievements of Computer
Science has been the development of abstract models of computation
that shield the discipline from the underlying technology. As
effective as these models have been, one must note, however, that they
\emph{embody several implicit physical assumptions}.  As Tommaso
Toffoli explains in his influential 1980 paper:

@@

Our logic of programs and our hardware are based on Boolean Logic. We go back
to Boole's 1853 book entitled \emph{An Investigation of the Laws of Thought,
  on which are Founded the Mathematical Theories of Logic and
  Probabilities}. The opening sentence of Ch.~1 is:
\begin{quote}
  The design of the following treatise is to investigate the fundamental laws
  of those operations of the mind by which reasoning is performed; \ldots
\end{quote}
A few chapters later, we find:
\begin{quote}
  \textbf{Proposition IV.}  That axiom of metaphysicians which is termed the
  principle of contradiction, and which affirms that it is impossible for any
  being to possess a quality, and at the same time not to possess it, is a
  consequence of the fundamental law of thought, whose expression is $x^2 =
  x$.
\end{quote}
A detailed historical analysis of Boole's ideas are beyond the scope of this
book. The above quotes, however, should convey the idea that our hardware is
based on assumptions that Boole found as reasonable laws in 1853.

Our theory of computation and complexity is based on the Turing Machine. We
go back to Turing's 1936 article entitled \emph{On Computable Numbers, with
  an Application to the Entscheidungsproblem,}. The opening sentence of
Sec. 1 is:
\begin{quote}
  We have said that the computable numbers are those whose decimals are
  calculable by finite means \ldots the justification lies in the fact that
  the human memory is necessarily limited.
\end{quote}
In Sec. 9, we find:
\begin{quote}
I think it is reasonable to suppose that they can only be squares
whose distance from the closest of the immediately previously observed
squares does not exceed a certain fixed amount.
\end{quote}
Again, a detailed historical analysis of Turing's ideas are beyond the scope
of this book. The above quotes should again convey the idea that our theories
of computation and complexity are based on some physical assumptions that
Turing found reasonable in 1936. When translated to physical language, the
assumptions embody what is known as the ``Bekenstein bound:'' which is an
upper limit on the amount of information that can be contained within a given
finite region of space.

One can examine more and more cases but the general story should be
clear. Our abstractions were made up based on a certain understanding of the
laws of physics. And since, as we argue, our abstractions are becoming at
odds with practice, it is time to revisit these abstractions. In the words of
Girard:
\begin{quote}
  In other terms, what is so good in logic that quantum physics should obey?
  Can't we imagine that our conceptions about logic are wrong, so wrong that
  they are unable to cope with the quantum miracle? 
  \\
  Instead of teaching logic to nature, it is more reasonable to learn
  from her. Instead of interpreting quantum into logic, we shall
  interpret logic into quantum (Girard 2007).
\end{quote}
Indeed one should take the physical principles underlying quantum mechanics,
the most successful physical theory known to us and adapt computation to
``learn'' from these principles. To illustrate the depth of our crisis, Scott
Aaronson, Umesh Vazirani, and others have proposed the following puzzle.

@@

Conventional classical models of computation, including boolean
logic, the Turing machine, and the $\lambda$-calculus, are founded on
primitives which correspond to \emph{irreversible} physical processes which
gratuitously erase information.  For example, a \emph{nand} gate is an
irreversible logical operation in the sense that its inputs cannot generally
be recovered from observing its output, and so is the operation of overriding
a cell on a Turing machine tape with a new symbol, and so is a
$\beta$-reduction which typically erases or duplicates values in a way that
is destructive and irreversible.

These irreversible abstractions chosen as the basis of our
models of computing are at ``odds'' with the underlying
reversible physical reality. The information theoretic underpinnings
of physics is best understood through the remarkable \emph{Landauer
  principle}~\cite{Landauer:1961}. Landauer, through his analysis of
Maxwell's Demon, established the idea that \emph{information is a
physical quantity} in much the sense that one would think of an
electron as a physical quantity, and that a certain minimum amount of
thermodynamic work has to be carried out to erase one bit of
information. Experimental verification of the Landauer principle
happened in recent years~\cite{berut2012experimental}. If information
has a physical significance, then it is subject to conservation laws
as mass and energy are.

@@

The Landauer Principle states that erasing information requires
energy. In other words, the password checking function must have
erased 1.2 bits during its calculation which dissipate as heat in any
physical realization of the function. In conventional models of
computation, this erasure is \emph{implicit}, i.e., it occurs as a
side effect of calculation. A big part of our development will be
concerned with making that erasure explicit.

The discussion in this section suggests that our foundations are really in
need of revision. There are many principles we can embrace.  We will start
with one, ``\textbf{conservation of information}'' and follow its
consequences, which will turn out to be far reaching.

>>>>

\section{Models of Computation}
%%%

But first, we need to better understand our current models of
computation, as a prelude to understand where they lead us astray.

\subsection{Origins}

Aspects of computation are
quite old. M\"{u}ller (1786) first conceived of the idea of a
``difference machine'', which Babbage (1819--1822) was able to
construct. There are other computer precursors as well -- the first
stored programs were actually for looms, most notably those of
Bouchon (1725) which operated on a paper tape, and Jacquard (1804)
which operated by chains of punched cards.

But it was Alan Turing's seminal work in 1936 which established the idea
that computation has a formal interpretation and that all
computability can be captured within a formal system. Implicit in this
achievement however is the idea that abstract models of computation
are just that -- \emph{abstractions of computation realized in the
physical world.}  In fact, one of the major achievements of Computer
Science has been the development of abstract models of computation
that shield the discipline from the underlying technology. As
effective as these models have been, one must note, however, that they
\emph{embody several implicit physical assumptions}.  As Tommaso
Toffoli explains in his influential 1980 paper:

@@

Our logic of programs and our hardware are based on Boolean Logic
-- which really
ought to be called \emph{Piercean logic}, as Boole had a rather different
conception of logic which is rather far from our current models.
Nevertheless, going back
to Boole's 1853 book entitled \emph{An Investigation of the Laws of Thought,
  on which are Founded the Mathematical Theories of Logic and
  Probabilities}. The opening sentence of Ch.~1 is:
\begin{quote}
  The design of the following treatise is to investigate the fundamental laws
  of those operations of the mind by which reasoning is performed; \ldots
\end{quote}
A few chapters later, we find:
\begin{quote}
  \textbf{Proposition IV.}  That axiom of metaphysicians which is termed the
  principle of contradiction, and which affirms that it is impossible for any
  being to possess a quality, and at the same time not to possess it, is a
  consequence of the fundamental law of thought, whose expression is $x^2 =
  x$.
\end{quote}
A detailed historical analysis of Boole's ideas are beyond our scope.
The above quotes, however, should convey the idea that our notions of
computation generally date back to ideas that were thought reasonable
before \tilde 1900.

It is conventional to base
the theory of computation and complexity on the Turing Machine.
Going back to Turing's 1936 article \emph{On Computable Numbers, with
  an Application to the Entscheidungsproblem,}, the opening sentence of
Sec. 1 is:
\begin{quote}
  We have said that the computable numbers are those whose decimals are
  calculable by finite means \ldots the justification lies in the fact that
  the human memory is necessarily limited.
\end{quote}
In Sec. 9, we find:
\begin{quote}
I think it is reasonable to suppose that they can only be squares
whose distance from the closest of the immediately previously observed
squares does not exceed a certain fixed amount.
\end{quote}
A detailed historical account is again beyond our scope. The quotes above
should convey the ideas that our theories
of computation and complexity are based on some assumptions that
Turing found reasonable in 1936. Though it is worth noting that these assumptions
are both physical (on distances) and metaphysical (on restrictions of the
mind).  If we take the human mind to be a physical ``machine'' which does
computation, then when both of the above assumptions are translated to the
language of physics, they embody what is known as the ``Bekenstein bound:''
which is an upper limit on the amount of information that can be contained
within a given finite region of space.

One can examine more and more cases but the general story should be
clear. Our abstractions were made up based on a certain understanding of the
laws of physics. In particular, our understanding of physics has evolved
tremendously since 1900!  Thus it is time to revisit these abstractions,
especially with respect to quantum mechanics.
In the words of Girard:
\begin{quote}
  In other terms, what is so good in logic that quantum physics should obey?
  Can't we imagine that our conceptions about logic are wrong, so wrong that
  they are unable to cope with the quantum miracle?
  \\
  Instead of teaching logic to nature, it is more reasonable to learn
  from her. Instead of interpreting quantum into logic, we shall
  interpret logic into quantum (Girard 2007).
\end{quote}

\begin{comment}
\jc{The reason I commented this out is that it is under-justified. The
reader will simply not understand what these next few lines are really
saying.}
Indeed one should take the physical principles underlying quantum mechanics,
the most successful physical theory known to us and adapt computation to
``learn'' from these principles. To illustrate the depth of our crisis, Scott
Aaronson, Umesh Vazirani, and others have proposed the following puzzle.

@@

\subsection{From classical to quantum}

Conventional classical models of computation, including boolean
circuits, the Turing machine, and the $\lambda$-calculus, are founded on
primitives which correspond to \emph{irreversible} physical processes which
gratuitously erase information.  For example, a \emph{nand} gate is an
irreversible logical operation in the sense that its inputs cannot generally
be recovered from observing its output, and so is the operation of overriding
a cell on a Turing machine tape with a new symbol, and so is a
$\beta$-reduction which typically erases or duplicates values in a way that
is destructive and irreversible.

These irreversible abstractions chosen as the basis of our
models of computing are at odds with the underlying
reversible physical reality. The information theoretic underpinnings
of physics is best understood through \emph{Landauer's
principle}~\cite{Landauer:1961}. As mentionned previously in the
introduction, Landauer, through his analysis of
Maxwell's Demon, established the idea that \emph{information is a
physical quantity} in much the sense that one would think of an
electron as a physical quantity. More precisely, his principle implies
that a certain minimum amount of
thermodynamic work has to be carried out to erase one bit of
information. Experimental verification of the Landauer principle
happened in recent years~\cite{berut2012experimental}. If information
has a physical significance, then it is subject to conservation laws
just as mass and energy are.

\subsection{conservation of information}

@@

The Landauer Principle states that erasing information requires
energy. In other words, the password checking function must have
erased 1.2 bits during its calculation, which dissipate as heat in any
physical realization of the function. In conventional models of
computation, this erasure is \emph{implicit}, i.e., it occurs as a
side effect of calculation. We wish to look at models where
erasure is simply not allowed.

There are, in fact, many different principles of physics which are
at odds with our current models of computation. Even within quantum
mechanics itself, there are other ideas (such as superposition or
entanglement) which are not present in current models. We will however
restrict ourselves to exploring a single dimension in which our
foundations should be revised: \textbf{conservation of information}.
We will follow its consequences, which will turn out to be far reaching.
